---
title: "Exploring parameter conversions between the SIR model and the KM approximation"
---

Here is the SIR model: 
$$
\begin{aligned}
\frac{dS}{dt} &= - \beta S I\\
\frac{dI}{dt} &= \beta S I - \gamma I\\
\frac{dR}{dt} &= \gamma I
\end{aligned}
$$
Here's the approximation:
$$
\frac{dR}{dt} \approx a \,\textrm{sech}^2{(\omega\,t - \phi)}
$$
We want to be able to convert the SIR parameters to KM parameters (and also from KM parameters to SIR parameters).

Based on Bacaër, we have:
$$
\begin{aligned}
a &= \frac{\gamma^3 Q^2}{2 S_0 \beta^2}\\
\omega &= \frac{Q \gamma}{2}\\
\phi &= \textrm{arctanh}\left(\frac{\frac{\beta S_0}{\gamma} - 1}{Q} \right)\\
Q &= \sqrt{\left(\frac{\beta S_0}{\gamma} - 1 \right)^2 + 2 S_0 I_0 \frac{\beta^2}{\gamma^2}}
\end{aligned}
$$
Now, rewriting $R_e = \beta S_0/\gamma$, we have:
$$
\begin{aligned}
a &= \frac{2 \omega^2 S_0}{\gamma R_e^2}\\
\omega &= \frac{Q \gamma}{2}\\
\phi &= \textrm{arctanh}\left(\frac{R_e - 1}{Q} \right)\\
Q &= \sqrt{\left(R_e - 1 \right)^2 + 2 R_e^2 I_0/S_0}
\end{aligned}
$$
For example, let's assume:
```{r}
Re <- 5
gamma <- 1
S0 <- 1998
I0 <- 2
```
Let's convert to KM paramters:

```{r}
Q <- sqrt((Re-1)^2 + 2 * Re^2 * I0/S0)
phi <- atanh((Re-1)/Q)
omega <- (Q * gamma)/2
a <- 2 * omega^2 * S0/(gamma * Re^2)
print(c(Q=Q, phi=phi, omega=omega, a=a))
```

Now, let's try to convert back. In particular, we have:
$$
R_e = \frac{1}{2}\bigg( 1 + \sqrt{1 + \frac{4\,\omega\,I_0 \sinh\!{(2\phi)}}{a}}\bigg) 
$$
But this doesn't work!
```{r}
(1 + sqrt(1 + 4 * omega * I0 * sinh(2 * phi)/a))/2
```
Is there something wrong with the original derivation? Let's try again...

First, we start with:
$$
\begin{aligned}
Q &= \sqrt{\left(R_e - 1 \right)^2 + 2 R_e^2 I_0/S_0}\\
Q^2 &= \left(R_e - 1 \right)^2 + 2 R_e^2 I_0/S_0\\
Q^2 - \left(R_e - 1 \right)^2 &= 2 R_e^2 I_0/S_0\\
S_0 &= \frac{2 R_e^2 I_0}{Q^2 - \left(R_e - 1 \right)^2}\\
&= \frac{2 R_e^2 I_0}{((R_e-1)/\tanh(\phi) )^2 - \left(R_e - 1 \right)^2}\\
&= \frac{2 R_e^2 I_0}{\left(R_e - 1 \right)^2 \left(\frac{1}{\tanh^2(\phi)} - 1\right)}\\
&= \frac{2 R_e^2 I_0 \sinh^2(\phi)}{\left(R_e - 1 \right)^2}
\end{aligned}
$$
OK, I think I found a typo with the original Baecer et al article at this point. They use
$$
Q = \sqrt{\left(R_e - 1 \right)^2 + 2 R_e I_0/S_0}
$$
during the derivation which is missing a square in the second term!!!! Let's keep going with the derivation..

Since $\gamma = 2 \omega/Q$, we have:
$$
\begin{aligned}
a &= \frac{2 \omega^2 S_0}{\gamma R_e^2}\\
S_0 &= \frac{a \gamma R_e^2}{2 \omega^2}\\
&= \frac{a R_e^2 }{Q \omega}\\
&= \frac{a R_e^2 \tanh(\phi)}{\omega (R_e - 1)}\\
&= \frac{a R_e^2 \sinh(\phi)}{\omega (R_e - 1) \cosh(\phi)}
\end{aligned}
$$
Now, combining the two equations:
$$
\begin{aligned}
\frac{a R_e^2 \sinh(\phi)}{\omega (R_e - 1) \cosh(\phi)} &= \frac{2 R_e^2 I_0 \sinh^2(\phi)}{\left(R_e - 1 \right)^2}\\
\frac{a}{\omega \cosh(\phi)} &= \frac{2 I_0 \sinh(\phi)}{\left(R_e - 1 \right)}\\
R_e &= 1 + 2 \omega I_0 \sinh(\phi) \cosh(\phi)/a
\end{aligned}
$$
Let's test:
```{r}
1 + 2 * omega * I0 * sinh(phi) * cosh(phi)/a
```
This works.
